---
title: "Regression Models for Time-to-Event Data"
author: "Matthew Pawley"
date: "January 2021"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE)

library("flexsurv")
library("survival")
library("knitr")
library("dplyr")

cols = c("#E495A5", "#86B875", "#7DB0DD", "#FF8C00")

set.seed(123)
```

# The APGW distribution

## APGW as a custom distribution in flexsurv

```{r}
# APGW
hapgw <- function(x, phi, lambda, gamma, kappa, log=FALSE){
  lambda*gamma*phi^gamma*(x)^{gamma-1}*(1+((phi*x)^gamma)/(kappa+1))^(kappa-1)
}
Hapgw <- function(x, phi, lambda, gamma, kappa, log=FALSE){
  lambda*(kappa+1)/kappa*((1+(phi*x)^gamma/(kappa+1))^kappa-1)
}
kappa_trans <- function(x){log(x+1)}
kappa_invtrans<- function(x){exp(x)-1}

custom.apgw <- list(
  name="apgw",
  pars=c("phi","lambda","gamma","kappa"),
  location="lambda",
  transforms=c(log,log,log,kappa_trans),
  inv.transforms=c(exp,exp,exp,kappa_invtrans),
  inits=function(t){c(1,median(t)/log(2),1,1)})
```

```{r}
# APGW with scale parameter
hapgwscale<- function(x, lambda, gamma, kappa, theta, log=FALSE){
   hapgw(x, theta, lambda, gamma, kappa, log=FALSE)
}
Hapgwscale <- function(x, lambda, gamma, kappa, theta, log=FALSE){
   Hapgw(x, theta, lambda, gamma, kappa, log=FALSE)
}
custom.apgwscale <- list(
  name="apgwscale",
  pars=c("lambda","gamma","kappa","theta"),
  location="theta",
  transforms=c(log,log,kappa_trans,log),
  inv.transforms=c(exp,exp,kappa_invtrans,exp),
  inits=function(t) c(1,1,1,1)
)
```


```{r}
# APGW with frailty parameter
hapgwfrailty <- function(x, phi, gamma, kappa, theta, log=FALSE){
  hapgw(x, phi, theta, gamma, kappa, log=FALSE)
}
Hapgwfrailty <- function(x, phi, gamma, kappa, theta, log=FALSE){
  Hapgw(x, phi, theta, gamma, kappa, log=FALSE)
}
custom.apgwfrailty <- list(
  name="apgwfrailty",
  pars=c("phi","gamma","kappa","theta"),
  location="theta",
  transforms=c(log,log,kappa_trans,log),
  inv.transforms=c(exp,exp,kappa_invtrans,exp),
  inits=function(t) c(1,1,1,median(t)/log(2))
)
```

```{r}
# APGW with tilt parameter
hapgwtilt <- function(x, phi, lambda, gamma, kappa, theta, log=FALSE){
  z <- Hapgw(x, phi, lambda, gamma, kappa)
  num <- theta * hapgw(x, phi, lambda, gamma, kappa) * exp(z)
  den <- exp(H_lleg(z, theta))
  y <- num/den
  return(y)
}
Hapgwtilt <- function(x, phi, lambda, gamma, kappa, theta, log=FALSE){
  y <- Hapgw(x, phi, lambda, gamma, kappa, log=FALSE)
  y <- H_lleg(y, theta)
  return(y)
}
custom.apgwtilt <- list(
  name="apgwtilt",
  pars=c("phi","lambda","gamma","kappa","theta"),
  location="theta",
  transforms=c(log,log,log,kappa_trans,log),
  inv.transforms=c(exp,exp,exp,kappa_invtrans,exp),
  inits=function(t) c(1,median(t)/log(2),1,1,1)
)
```

```{r echo=TRUE}
# APGW with reverse-tilt parameter
hapgwrevtilt <- function(x, phi, lambda, gamma, kappa, theta, log=FALSE){
  z <- exp(H_lleg(x, theta))
  C <- theta*exp(x)/z
  y <- C*hapgw(log(z),phi,lambda,gamma,kappa)
  return(y)
}
Hapgwrevtilt <- function(x, phi, lambda, gamma, kappa, theta, log=FALSE){
  y <- Hapgw(H_lleg(x, theta), phi, lambda, gamma, kappa)
  return(y)
}
custom.apgwrevtilt <- list(
  name="apgwrevtilt",
  pars=c("phi","lambda","gamma","kappa","theta"),
  location="theta",
  transforms=c(log,log,log,kappa_trans,log),
  inv.transforms=c(exp,exp,exp,kappa_invtrans,exp),
  inits=function(t) c(1, median(t)/log(2), 1, 1, 1)
)
```

## Sampling from the APGW distribution

```{r echo=TRUE}
rapgw <- function(n, phi, lambda, gamma, kappa){
  v <- rexp(n=n, rate=1)
  x <- invHapgw(v, phi, lambda, gamma, kappa, log=FALSE)
  return(x)
}

rapgw_scale <- function(n, phi, lambda, gamma, kappa, theta){
  rapgw(n, theta*phi, lambda, gamma, kappa)
}

rapgw_frailty <- function(n, phi, lambda, gamma, kappa, theta){
  rapgw(n, phi, theta*lambda, gamma, kappa)
}

invHapgw <- function(x, phi, lambda, gamma, kappa, log=FALSE){
  (1/phi)*((kappa+1)*((1+kappa*x/(lambda*(kappa+1)))^(1/kappa)-1))^(1/gamma)
}

H_lleg <- function(x, theta){log(1+theta*(exp(x)-1))}

rapgw_tilt <- function(n, phi, lambda, gamma, kappa, theta){
  v <- rexp(n=n, rate=1)
  x <- H_lleg(v, 1/theta)
  x <- invHapgw(x, phi, lambda, gamma, kappa, log=FALSE)
  return(x)
}

rapgw_revtilt <- function(n, phi, lambda, gamma, kappa, theta){
  x <- rapgw(n, phi, lambda, gamma, kappa)
  x <- H_lleg(x, 1/theta)
  return(x)
}
```

## Testing the samplers and custom distributions

* Generate $N=5000$ samples from $\mathrm{APGW}(\phi=1.5,\lambda=1,\gamma=2,\kappa=-0.2)$ with no censoring.
* Plot sample histogram and true density function.
* Fit an APGW distribution to the data. Compare the parameter estimates to the true values. 

```{r}
# Generate APGW samples
n <- 5000
phi <- 1
lambda <- 5
gamma <- 2
kappa <- -0.2

t <- rapgw(n, phi, lambda, gamma, kappa)
delta <- rep(1, n)
df <- data.frame(t, delta)
```

```{r}
# Compare sample hist and true density function
fun <- function(x){
  hapgw(x, phi, lambda, gamma, kappa, log=FALSE) * exp(-Hapgw(x, phi, lambda, gamma, kappa, log=FALSE))
}
d <- density(t)
plot(d, main = NA, xlab = "t", ylab = "Density", col=cols[1], lwd=2)
curve(fun, add=TRUE, col=cols[2], lwd=2, lty=1, main=NA)
legend("topright", col = c(cols[1], cols[2]), bty = "n", lwd = 2, lty=c(1,1), 
       c("Kernel density estimate", "Actual"))
```


```{r message=FALSE}
# Fit APGW model and check estimates
fs1 <- flexsurvreg(Surv(t,delta)~1, data=df, dist=custom.apgw, fixedpars=1)
fs1
```

## Testing the APGW custom distribution on the bc dataset

```{r message=FALSE}
# Fit an APGW model
fs1 <- flexsurvreg(Surv(recyrs,censrec)~phi(group), 
                   data=bc, 
                   dist=custom.apgw, 
                   fixedpars=c(2,4),
                   inits=c(1,1,1,1))
fs1
```

```{r message=FALSE}
# Fit a Weibull model\
fs2 <- flexsurvreg(Surv(recyrs,censrec)~group, 
                   data=bc, 
                   dist="weibull")
fs2
```

# Simulation studies

## Simulate data in groups

```{r}
sim_df <- function(K, nk, t_max, c_rate, phi, lambda, gamma, kappa, theta_1, beta, family){

  group <- rep(1:K, nk)
  theta <- ifelse(group==1, theta_1, theta_1*exp(beta[group-1]))
  theta <- theta_1 * exp(c(rep(0,nk[1]), beta[group-1]))
  
  # simulate event times
  if(family == "scale"){t <- mapply(rapgw_scale, phi, lambda, gamma, kappa, theta, MoreArgs=list(n=1))}
  else if(family == "frailty"){t <- mapply(rapgw_frailty, phi, lambda, gamma, kappa, theta, MoreArgs=list(n=1))}
  else if(family == "tilt"){t <- mapply(rapgw_tilt, phi, lambda, gamma, kappa, theta, MoreArgs=list(n=1))}
  else if(family == "revtilt"){t <- mapply(rapgw_revtilt, phi, lambda, gamma, kappa, theta, MoreArgs=list(n=1))}
  else stop(cat("family must be \"scale\", \"frailty\", \"tilt\", or \"revtilt\""))

  # censoring 
  c <- rexp(n=sum(nk), rate=c_rate)
  c <- pmin(c, t_max)
  delta <- 1*(t<c)
  t <- pmin(t,c)
  
  # create dataframe
  df <- data.frame("group"=as.factor(group), t, delta)
  
  return(df)
}
```

## Experiment: frailty (PH) data

* Simulate PH data with 3 groups
* Make diagnostic plots for AFT/PH/PO
* Fit AFT/PH/PO/PGT models
* Compare CH functions of PH and PO models
* Test the effect of varying the censoring mechanism

```{r}
# Simulate data
K <- 3
nk <- rep(300,K)
t_max <- 1.5
c_rate <- 0.1
phi <- 1
lambda <- 1
gamma <- 4
kappa <- -0.2
theta_1 <- 2
beta <- c(0.05, 0.6)

df <- sim_df(K, nk, t_max, c_rate, phi, lambda, gamma, kappa, theta_1, beta, "frailty")
```

```{r}
# Plot the survival curves
km <- survfit(Surv(t, delta) ~ group, data = df)
plot(km, lty = 1:K, xlab = "t",  ylab = "S(t)") 
legend("bottomleft", lty = 1:K, bty = "n", paste("Group",1:K))
```

```{r}
# Check PH assumption - Schoenfeld residuals test
res.cox <- coxph(Surv(t, delta) ~ group, data=df)
test.ph <- cox.zph(res.cox)
test.ph
```

```{r}
# Check PH assumption - plot Schoenfeld residuals
plot(test.ph)
```

```{r}
# Check PH assumption - plot log-log survival curves
plot(km, fun=function(x) {log(-log(x))},
     lty = 1:K, xlab = "t",  ylab = "log(-log S(t))", log="x") 
legend("topleft", lty = 1:K, bty = "n", paste("Group",1:K))
```

```{r}
# Check AFT assumption - plot log-time survival curves
km_log <- survfit(Surv(log(t), delta) ~ group, data = df)
plot(km_log, fun="S",
     lty = 1:K, xlab = "log(t)",  ylab = "S(t)") 
legend("topleft", lty = 1:K, bty = "n", paste("Group",1:K))
```

```{r}
# Check AFT assumption - Bartlett test
g1 <- df$t[df$group==1 & df$delta==1]
g2 <- df$t[df$group==2 & df$delta==1]
g3 <- df$t[df$group==3 & df$delta==1]
k <- 3
n1 <- length(g1)
n2 <- length(g2)
n3 <- length(g3)
N <- n1 + n2 + n3
var1 <- var(log(g1))
var2 <- var(log(g2))
var3 <- var(log(g3))
Sp2 <- ((n1-1)*var1 + (n2-1)*var2 + (n3-1)*var3) / (N-k)
test_stat <- ((N-k)*log(Sp2) - ((n1-1)*log(var1) + (n2-1)*log(var2) + (n3-1)*log(var3)))/(1+(1/(n1-1)+1/(n2-1)+1/(n3-1)-1/(N-k))/(3*(k-1)))
p_value <- pchisq(test_stat, df=k-1, lower.tail = FALSE)
```

```{r echo=FALSE}
print(paste("The sample variances of the log survival times for each group are", round(var1,3),",", round(var2,3),"and", round(var3,3), "respectively."))
print(paste("Bartlett test p-value:", round(p_value,3)))
```

```{r}
# Check PO assumption - log odds curves
plot(km, fun=function(x) {log((1-x)/x)},
     lty = 1:K, xlab = "t",  ylab = "Log odds", log="x") 
legend("topleft", lty = 1:K, bty = "n", paste("Group",1:K))
```

```{r message=FALSE}
# Fit AFT/PH/PO/PGT models
fs_aft <- flexsurvreg(Surv(t,delta)~group, data=df, dist=custom.apgwscale, fixedpars=1)
fs_ph <- flexsurvreg(Surv(t,delta)~group, data=df,  dist=custom.apgwfrailty, fixedpars=1)
fs_po <- flexsurvreg(Surv(t,delta)~group, data=df, dist=custom.apgwtilt, fixedpars=1)
fs_pgt <- flexsurvreg(Surv(t,delta)~group, data=df, dist=custom.apgwrevtilt, fixedpars=1)
```

```{r}
# Compare the CH functions of the PH and PO 
plot(fs_ph, col = cols[1], lwd.obs = 1, lty.obs=2, type="cumhaz", xlab = "t", ylab = "H(t)", yaxs="i", xaxs="i")
lines(fs_po, col = cols[2], lwd.obs = 1, lty.obs=1, type="cumhaz")
legend("topleft", col = c("black", cols[1], cols[2]),  lty = c(2,1,1), bty = "n", lwd = 1, c("Non-parametric estimate", "APGW PH", "APGW PO"))
```

```{r eval=FALSE, message=FALSE, echo=TRUE}
# Test effect of varying censoring mechanism - this is intensive so not run here
K <- 3
nk <- rep(300,K)
c_rate <- 0.0001
phi <- 1
lambda <- 1
gamma <- 4
kappa <- -0.2
theta_1 <- 2
beta <- c(0.05, 0.6)

t_max_vals <- seq(from=0.5, to=2.5, by=0.2)
M <- 20
ph_aic <- matrix(NA, nrow=length(t_max_vals), ncol=M)
po_aic <- matrix(NA, nrow=length(t_max_vals), ncol=M)

for (i in 1:length(t_max_vals)){
  t_max <- t_max_vals[i]
  for (j in 1:M){
    df <- sim_df(K, nk, t_max, c_rate, phi, lambda, gamma, kappa, theta_1, beta, "frailty")
    tryCatch(
      expr = {
        fs_ph <- flexsurvreg(Surv(t,delta)~group, data=df, dist=custom.apgwfrailty, fixedpars=1)
        ph_aic[i,j] <- fs_ph$AIC
      },
      error = function(e){
        ph_aic[i,j] <- NA
      }
    )
    tryCatch(
      expr = {
        fs_po <- flexsurvreg(Surv(t,delta)~group, data=df, dist=custom.apgwtilt, fixedpars=1)
        po_aic[i,j] <- fs_po$AIC
      },
      error = function(e){
        po_aic[i,j] <- NA
      }
    )
  }
}

ph_aic <- rowMeans(ph_aic, na.rm = TRUE)
po_aic <- rowMeans(po_aic, na.rm = TRUE)
plot(t_max_vals, po_aic-ph_aic, type="l", 
     xlab = "Administrative censoring time",
     ylab = "Difference in AICs of PH and PO models")
```

# Fibres data

* Plot diagnostics
* Fit PO models

```{r}
# Load data
fibres_df <- read.csv(file = 'fibres_failure_stress.csv', header = TRUE)
fibres_df$length <- as.factor(fibres_df$length)
fibres_df$delta <- 1
```

```{r}
# Plot survival curves
km <- survfit(Surv(fail_stress, delta) ~ length, data = fibres_df)
plot(km, lty = 1:4, xlab = "t",  ylab = "S(t)") 
legend("bottomleft", lty = 1:4, bty = "n", c("1mm","10mm","20mm","50mm"))
```

```{r}
# Check PH assumption - Schoenfeld residuals test
res.cox <- coxph(Surv(fail_stress, delta) ~ length, data = fibres_df)
test.ph <- cox.zph(res.cox)
test.ph
```

```{r}
# Check PH assumption - plot Schoenfeld residuals
plot(test.ph)
```

```{r}
# Check PH assumption - plot log-log survival curves
plot(km, fun=function(x) {log(-log(x))}, lty = 1:4, xlab = "t",  ylab = "log(-log S(t))", log="x") 
legend("topleft", lty = 1:4, bty = "n", c("1mm","10mm","20mm","50mm"))
```

```{r}
# Check AFT assumption - plot log-time survival curves
km_log <- survfit(Surv(log(fail_stress), delta) ~ length, data = fibres_df)
plot(km_log, lty = 1:4, xlab = "log(t)",  ylab = "S(t)") 
legend("bottomleft", lty = 1:4, bty = "n", c("1mm","10mm","20mm","50mm"))
```

```{r}
# Check PO assumption - log-odds curves
plot(km, fun=function(x) {log((1-x)/x)}, lty = 1:4, xlab = "t",  ylab = "Log odds", log="x") 
legend("topleft", lty = 1:4, bty = "n", c("1mm","10mm","20mm","50mm"))
```


```{r}
# Fit AFT model
fs <- flexsurvreg(Surv(fail_stress, delta) ~ length, data=fibres_df, dist=custom.apgwscale, fixedpars=1)
fs
```

# PBC dataset

* Pre-processing
* Categorise patients into groups
* Fit PO model

```{r}
# pre-processing
pbc$time <- pbc$time / 365
pbc$status[pbc$status==1] <- 0
pbc$status[pbc$status==2] <- 1
pbc$logalbumin <- log(pbc$albumin) - mean(log(pbc$albumin))
pbc$logbili <- log(pbc$bili) - mean(log(pbc$bili))
pbc$protime[is.na(pbc$protime)] <- mean(pbc$protime, na.rm = TRUE)
pbc$logprotime <- log(pbc$protime) - mean(log(pbc$protime))

kable(head(pbc))
```

```{r fig.height=10}
# Plot histograms to split patients into groups
# Check PO assumption for these groups

par(mfrow=c(2,2))

hist(pbc[,"logalbumin"], xlab="logalbumin", main=NULL)
hist(pbc[,"logprotime"], xlab="logprotime", main=NULL)

pbc$logalbumincat <- cut(pbc$logalbumin, breaks=c(-Inf, -0.05, Inf),   labels=c("Low","High"))  

km_logalb <- survfit(Surv(time, status) ~ logalbumincat, data = pbc)
plot(km_logalb, fun=function(x) {log((1-x)/x)}, lty = 1:2, xlab = "t",  ylab = "Log odds", log="x") 
legend("topleft", lty = 1:2, bty = "n", c("Low","High"),title="log(albumin)")

pbc$logprotimecat <- cut(pbc$logprotime, breaks=c(-Inf, 0, Inf),labels=c("Low","High"))  

km_logprotime <- survfit(Surv(time, status) ~ logprotimecat, data = pbc)
plot(km_logprotime , fun=function(x) {log((1-x)/x)}, lty = 1:2, xlab = "t",  ylab = "Log odds", log="x") 
legend("topleft", lty = 1:2, bty = "n", c("Low","High"), title="log(protime)")
```


```{r message=FALSE}
# Fit PO model
fs_po <- flexsurvreg(Surv(time, status) ~ age + edema + logbili + logalbumin + logprotime, data=pbc, dist=custom.apgwtilt, inits=c(1/18.4,1,1.54,1,1), fixedpars=2)
fs_po
```


